Road User video evidence of road traffic offences: Preliminary analysis of Operation Snap data and suggestions for a research agenda

### 2024-07-25 paper.qmd
Updates...

There was an increase in submissions to opsnap over time between 2021 and 2023.
This may reflect changing levels of road use during and after the Covid-19 pandemic, increased awareness of OpSnap, or increased video camera ownership levels.
Comparative analysis including other police jurisdictions should be undertaken to determine whether this experience is common or isolated.
We hypothesise that it is a common experience and also that that the volume of submissions will continue to increase in the near and middle term (though ideally, a decline would occur as roads become safer).

If submission volumes continue to increase, means of informing triage and prioritisation of cases will be increasingly valuable.
Triage systems are probably already in use in different areas, and research to identify and inform best practice should be undertaken.
Likewise, research to determine best practice in determining disposal recommendations will prove fruitful.
It may provide feasible, perhaps using machine learning, to develop automated means of triage, determination of offence type and disposal.
Research to gauge the differential effectiveness of disposal measures is needed, starting with measures of recidivism.

Within the trend of increasing video submissions there were distinct seasonal patterns.
Seasonal variation in cyclists' submission reflect seasonal (weather and other) influences on cycling prevalence and duration.
This suggests seasonal variation in preventive responses might be tailored to need.
It is well established that crime is highly concentrated along whatever dimension is examined [@farrellPreventingRepeatRepeat2017].
With respect to video evidence this will include recidivists (repeat offenders), repeat submitters of evidence (who may or may not be repeat victims/survivors), the concentration of incidents (close passes, near misses and crashes) at certain times and places, with different types of experiences concentrated among certain types of road users.
For example, we hypothesise that close passes are more likely to be reported by cyclists and horse riders than vehicle drivers, both due to the nature of the road user interactions and the relative risk associated with the manoeuvre.
As such, the types of incidents reported by vehicle drivers would be systematically different, indicating a need for different investigative and preventive approaches.
Policing and preventions efforts focused on where problems are concentrated are more resource efficient -- it is the foundation of problem-oriented policing (see e.g., (@laingAggressiveDriving2010; @scottSpeedingResidentialAreas2010)- and research using video submissions holds the potential to analysis of problem concentrations.

Studies focused on specific types of road user will prove informative with respect to policy and practice.
Dash-cam submissions by motorists offers great potential to inform other aspects of road safety.
Research into submissions by pedestrians, horse riders and motorcyclists should prove viable to improve the safety of these parties.
There were over four hundred submissions by horse riders in West Yorkshire, which means there will be thousands nationwide.
The British Horse Society[^1] and others may be interested in this data being used to inform the safety of horses and riders.
The British Motorcyclists Federation[^2] may be interested in research promoting the safety of its members and in why motorcyclist submissions are relatively infrequent.

>>>>

There has been an increase in submissions to West Yorkshire Police's OpSnap system since it was set-up in 2021.
This may reflect changing levels of road use during and after the Covid-19 pandemic, increased awareness of OpSnap, or increased video camera ownership levels.
Comparative analysis including other police jurisdictions should be undertaken to determine whether this experience is common or isolated.
We hypothesise that it is a common experience for police forces (and other organisations) setting up public video submissions reporting antisocial and dangerous driving and that that the rate of submissions will continue to increase (longer term, ideally, submission rates decline as roads become safer).

This was a case study of West Yorkshire but each of the 43 police forces in England and Wales collates OpSnap data.
Our preliminary scoping of individual force websites suggests OpSnap data is publicly available nationwide.
A national dataset should be developed that includes OpSnap data from all police forces.
 A feasibility study should be undertaken to establish cross-force data availability and compatibility.
A national dataset holds potential for national-level, cross-regional, and comparative analyses of patterns and trends.
It would have the potential to promote cross-national comparative analysis and international cooperation in road safety should similar data be collated elsewhere.
It would hold the potential for the development of rankings according to different criteria and, thereby, potential performance-related metrics [@tiwana2015].
There will be added-value from a national dataset when multiple incidents involve the same persons in different jursdictions, such as recidivist drivers, and in relation to other cross-jurisdictional issues.

With submission volumes likely to continue to increase, means of informing triage and prioritisation of cases will be increasingly valuable.
Triage systems are almost certainly used already, and research to identify best practice should be undertaken.
Similarly, research to determine best practice in determining disposal recommendations should be undertaken.
It may provide feasible, perhaps using machine learning, to develop automated means of triage, determination of offence type and disposal.
Research to gauge the differential effectiveness of disposal measures is also needed.
A starting point for an evaluation may be easures of recidivism found in the OpSnap data.

Within the trend of increasing video submissions there were distinct seasonal patterns.
Seasonal variation in submissions reflect seasonal (weather and other) influences, and these will vary by transport mode: cycling, for example, is less prevalent in winter.
This suggests seasonal variation in preventive responses might be tailored to need.

Future research should focus on different types of problem concentration.
It is well established that crime is highly concentrated along whatever dimension is examined [@farrellPreventingRepeatRepeat2017].
With respect to video evidence this will include recidivists (repeat offenders), repeat submitters of evidence (who may or may not be repeat victims/survivors), the concentration of incidents (close passes, near misses and crashes) at certain times and places, with different types of experiences concentrated among certain types of road users.
As discussed in the introduction, close passes are more likely to be reported by cyclists and horse riders than vehicle drivers, both due to the nature of the road user interactions and the relative risk associated with the maneuver.

There was preliminary confirmation of this pattern in our analysis which found that offence types reported by vehicle drivers are systematically different to those of other road users.
Hence future research should consider that investigative and preventive approaches should be tailored to different contexts and types of road user.
Rural roads account for a disproportionate amount of fatalities @brake, and further research using video evidence may inform preventive approaches tailored to road type.

Policing and prevention efforts focused on where problems are concentrated are more resource efficient — it is the foundation of problem-oriented policing (see e.g., (@laingAggressiveDriving2010; @scottSpeedingResidentialAreas2010).

Studies focused on specific types of road user will prove informative with respect to policy and practice.
Dash-cam submissions by motorists offers great potential to inform other aspects of road safety.
Research into submissions by pedestrians, horse riders and motorcyclists should prove viable to improve the safety of these parties.
There were over four hundred submissions by horse riders in West Yorkshire, which means there will be thousands nationwide.
The British Horse Society[^1] and others may be interested in this data being used to inform the safety of horses and riders.
The British Motorcyclists Federation[^2] may be interested in research promoting the safety of its members and in why motorcyclist submissions are relatively infrequent
.
